---
title: "Introduction to non-metric multidimensional scaling"
fontfamily: mathpazo
fontsize: 10pt
geometry: margin=0.75in
output:
  pdf_document:
    toc: no
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
layout: topic
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 5, fig.height = 3.5, 
                      echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

This is a brief tutorial on non-metric multidimensional scaling using the `vegan` package in `R`. I use data from an experiment that tested the effects of urchin grazing on the community structure of subtidal rock walls ([Elahi and Sebens 2012](https://doi.org/10.3354/meps09603)). 

### Load packages

```{r load-packages, message=FALSE, eval = TRUE}
library(vegan)
library(tidyverse)
theme_set(theme_bw(base_size = 14) + 
            theme(strip.background = element_blank(), 
                  panel.grid = element_blank()))

## Define the repository from which we get the data
repo_url <- "https://raw.githubusercontent.com/elahi/phd_elahi/master/"

```

### Load data

```{r load-data, eval = TRUE}
sjc_sessile <- read_csv(
  paste(repo_url,
        "Elahi_sessile_quadrat/output/sjc_master_compiled/sessile_compiled_final_wide.csv", 
        sep = ""))
dat <- sjc_sessile %>% 
  filter(dataset == "elahi_urchin_addition") %>% 
  select(-X1, -X)
names(dat)[1:10]
```

### Explore data

```{r explore-data}
dat %>% 
  count(site, urchin_addition, transect)
  
dat %>% count(dateR) 

dat <- dat %>% filter(dateR == "2009-09-24")

```

### Examine percent cover of all sessile taxa and functional groups

```{r}
datL <- dat %>% 
  gather(key = species, value = cover, bare_rock:rubble)

cover_means <- datL %>%
  group_by(species) %>% 
  summarise(mean = mean(cover)) %>% 
  arrange(mean)

cover_means %>% 
  ggplot(aes(mean)) + 
  geom_histogram(binwidth = 0.01) 
```

### Remove uncommon sessile taxa

Here, I use a 1% cover threshold. I do this for simplicity - not for biological relevance. 

```{r}
species_subset <- cover_means %>% filter(mean >= 0.01)
species_subset
```

### Prepare data for ordination

```{r prep-data}
# create spp matrix 
abun <- dat[, names(dat) %in% species_subset$species]

# transform matrix
abun_sq <- sqrt(abun)

# presence/absence
abun_pa <- decostand(abun, method="pa")

# create group matrix
dat_group <- dat[,1:8]
```

To illustrate the approach, I use the default settings in the function `metaMDS` to run the mds on proportional cover data (i.e., these data were not transformed). But later, I will run the mds on square-root transformed data, and presence/absence data. 

```{r trial}
## Rename the matrix
mds_mat <- abun

## Run nmds, using default settings
mds_results <- metaMDS(mds_mat)

# Using the scores function from vegan to extract the site scores and convert to a data.frame
data_scores <- as.data.frame(scores(mds_results))  

# Combining with group dataframe
data_scores <- cbind(data_scores, dat_group)
data_scores %>% head()

# Using the scores function from vegan to extract the species scores and convert to a data.frame
species_scores <- as.data.frame(scores(mds_results, "species"))  

# Create a column of species, from the rownames of species.scores
species_scores$species <- rownames(species_scores)  
head(species_scores)  #look at the data
```

```{r}
data_scores %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, shape = urchin_addition,colour = urchin_addition)) + 
  geom_point(size = 2, alpha = 0.8) + # add the point markers
  coord_equal() +
  theme(axis.text = element_blank(),  # remove axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title = element_blank(), # remove axis labels
        legend.title = element_blank()) 
```

Now separate by site. 

```{r, fig.width = 7, fig.height = 5}
data_scores %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, shape = urchin_addition,colour = urchin_addition)) + 
  geom_point(size = 2, alpha = 0.8) + # add the point markers
  coord_equal() +
  theme(axis.text = element_blank(),  # remove axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title = element_blank(), # remove axis labels
        legend.title = element_blank()) + 
  facet_wrap(~ site)# remove y-axis labels)
```

### Comparing NMDS after transformations

```{r compare-mds, results = 'hide'}

# Perform the NMDS in 2 dimensions on raw data
set.seed(132)
abun_MDS2 <- metaMDS(abun, distance="bray", k=2, engine = 'monoMDS', 
                    autotransform=FALSE, noshare=0.1, trymax=40, zerodist='add')

# Perform the NMDS in 2 dimensions on sqrt data
set.seed(132)
abun_sq_MDS2 <- metaMDS(abun_sq, distance="bray", k=2, engine = 'monoMDS', 
                       autotransform=FALSE, noshare=0.1, trymax=40, zerodist='add')

# Perform the NMDS in 2 dimensions on presence / absence data
set.seed(12345678) # was difficult to converge
abun_pa_MDS2 <- metaMDS(abun_pa, distance="bray", k=2, engine = 'monoMDS', 
                       autotransform=FALSE, noshare=0.1, trymax=40, zerodist='add')

```

```{r mds-functions}

extract_data_scores <- function(mds_results, dat_group){
# Using the scores function from vegan to extract the site scores and convert to a data.frame
data_scores <- as.data.frame(scores(mds_results))  
# Combining with group dataframe
data_scores <- cbind(data_scores, dat_group)
return(data_scores)
}

extract_species_scores <- function(mds_results){
# Using the scores function from vegan to extract the species scores and convert to a data.frame
species_scores <- as.data.frame(scores(mds_results, "species"))  
# Create a column of species, from the rownames of species.scores
species_scores$species <- rownames(species_scores)  
return(species_scores)
}

```

```{r compare-mds-extract}

abun_MDS2$stress
abun_sq_MDS2$stress 
abun_pa_MDS2$stress 


abun_ds <- extract_data_scores(mds_results = abun_MDS2, dat_group = dat_group) %>% 
  mutate(transformation = "none")
abun_sq_ds <- extract_data_scores(mds_results = abun_sq_MDS2, dat_group = dat_group) %>% 
  mutate(transformation = "square-root")
abun_pa_ds <- extract_data_scores(mds_results = abun_pa_MDS2, dat_group = dat_group) %>% 
  mutate(transformation = "presence")

data_scores <- rbind(abun_ds, abun_sq_ds, abun_pa_ds) %>% 
  mutate(transformation = factor(transformation, 
                                 levels = c("none", "square-root", "presence")))

```

```{r compare-mds-plot, fig.width = 7, fig.height = 7}
data_scores %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, shape = urchin_addition,colour = urchin_addition)) + 
  geom_point(size = 2, alpha = 0.8) + # add the point markers
  coord_equal() +
  theme(axis.text = element_blank(),  # remove axis text
        axis.ticks = element_blank(),  # remove axis ticks
        axis.title = element_blank(), # remove axis labels
        legend.title = element_blank()) + 
  facet_grid(transformation ~ site)# remove y-axis labels)
```


<!-- * * * -->

<!-- ## More Practice -->

<!-- 7.  Mutate the data frame so that it includes a new variable that contains the  -->
<!--     average speed, `avg_speed` traveled by the plane for each flight (in mph). -->
<!--     **Hint:** Average speed can be calculated as distance divided by -->
<!--     number of hours of travel, and note that `air_time` is given in minutes. -->
<!--     Make a scatterplot of `avg_speed` vs. `distance`. **Hint:** Use `geom_point()`.  -->
<!--     Describe the relationship between average speed and distance. -->

<!-- 8.  Replicate the following plot. **Hint:** The data frame plotted only -->
<!--     contains flights from American Airlines, Delta Airlines, and United -->
<!--     Airlines, and the points are `color`ed by `carrier`. Include the code you used to create the plot,  -->
<!--     as well as the plot itself.  -->
<!--     <!-- Once you replicate the plot, determine (roughly) what the cutoff point is for departure --> -->
<!--     <!-- delays where you can still expect to get to your destination on time. --> -->

<!-- ## License   -->
<!-- <div id="license"> -->
<!-- Robin Elahi, that is released under a [Creative Commons Attribution-ShareAlike 4.0](http://creativecommons.org/licenses/by-sa/4.0) license. -->
<!-- This lab was adapted from a lab written by.. -->
<!-- </div> -->

