---
title: "Introduction to R, RStudio, and Rmarkdown"
output:
  word_document: 
    highlight: pygments
editor_options:
  chunk_output_type: console
---

```{r include = FALSE, cache = FALSE}
# This code block sets up the r session when the page is rendered to html / word / pdf

# Change general options
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, include = TRUE,  
                      message = FALSE, warning = FALSE, 
                      cache = FALSE, par = FALSE)

# Set up figure options
knitr::opts_chunk$set(fig.align = 'center', fig.show = 'asis', 
                      fig.width = 3.5, fig.height = 3.5) # default fig sizes

```

### Present day birth records in the U.S.
In the previous few pages, you recreated some of the displays and preliminary 
analysis of Arbuthnot's baptism data. Your assignment involves repeating these 
steps, but for present day birth records in the United States.  Load up the 
present day data with the following command.

```{r load-present-data}
source("http://www.openintro.org/stat/data/present.R")
```

The data are stored in a data frame called `present`.

What years are included in this data set? What are the dimensions of the data frame and what are the variable or column names?

```{r}
summary(present)
dim(present)
```

> The years range from 1940 to 2002.  
> The dimensions are 63 rows and 3 columns.  
> The column names are year, boys, girls.  

How do these counts compare to Arbuthnot's? Are they on a similar scale?

> These counts are 3 orders of magnitude higher than Arbuthnot's. 

Make a plot that displays the boy-to-girl ratio for every year in the data 
    set. What do you see? Does Arbuthnot's observation about boys being born in
    greater proportion than girls hold up in the U.S.? Include the plot in your 
    response.

```{r}
bg_ratio <- with(present, boys / girls)
plot(present$year, bg_ratio, type = "l", xlab = "Year", ylab = "Boy:girl ratio")
```

> The ratio is above 1, meaning that more boys than girls are being born in the US - similar to Arbuthnot. In addition, there is a decreasing trend. 

In what year did we see the most total number of births in the U.S.? You can
    refer to the help files or the R reference card 
    [http://cran.r-project.org/doc/contrib/Short-refcard.pdf](http://cran.r-project.org/doc/contrib/Short-refcard.pdf)
    to find helpful commands.

```{r}
births <- with(present, boys + girls)
present$year[which.max(births)]
```

> The highest number of births occurred in 1961
