---
title: 'Inference for numerical data II'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
  editor_options:
    chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## American salaries

Since 2005, the American Community Survey polls ~$3.5 million households yearly. We will work 
with a random sample of 2000 observations from the 2012 ACS. Throughout the lab, you will begin to use the `dplyr` package for data wrangling and summarization. For more details, see [r4ds chapter on data transformation](https://r4ds.had.co.nz/transform.html). 

```{r load-data}
library(dplyr)
library(ggplot2)
load(url("https://stat.duke.edu/~mc301/data/acs.RData"))
```

```{r, eval = FALSE, echo = FALSE}
load("oiLabs-elahi/inf_for_numerical_data_2/acs.RData")
```

Below are the *metadata* for this dataset:

* `income`: Yearly income (wages and salaries)
* `employment`: Employment status, not in labor force, unemployed, or employed
* `hrs_work`: Weekly hours worked
* `race`: Race, White, Black, Asian, or other
* `age`: Age
* `gender`: gender, male or female
* `citizens`: Whether respondent is a US citizen or not
* `time_to_work`: Travel time to work
* `lang`: Language spoken at home, English or other
* `married`: Whether respondent is married or not
* `edu`: Education level, hs or lower, college, or grad
* `disability`: Whether respondent is disabled or not
* `birth_qrtr`: Quarter in which respondent is born, jan thru mar, apr thru jun, jul thru sep, or oct thru dec

Note that this dataset contains some people who are not in the labor force or not employed. 
First, let's subset the dataset for those who are employed. We will call this new dataset 
`acs_emp`, short for "employed". Here we use the `filter` function from the `dplyr` package for subsetting the data based on attributes stored in a variable.

```{r subset-employed}
acs_emp <- acs %>%
  filter(employment == "employed")
```

1.  What percent of the original sample (`acs`) are employed?

Next let's take a look at the income distribution by gender. The first step would be to
create a visualization:

```{r gender-income-viz}
ggplot(data = acs_emp, aes(x = gender, y = income)) + 
  geom_boxplot()
```

We can also obtain summary statistics such as means, standard deviations, and sample sizes.

```{r gender-income-summary}
acs_emp %>%
  group_by(gender) %>%
  summarise(xbar = mean(income), 
            s = sd(income), 
            n = n())
```

1.  At a first glance how do the average incomes of males and females compare? Make sure
    to include the visualization and the summary statistics in your answer, and discuss/
    interpret them.

Before you proceed, make sure to load the `inference` function. 

```{r load-inference}
load(url("https://stat.duke.edu/~mc301/R/fun/inference.RData"))
```
    
1.  Construct a 95% confidence interval for the difference between the average incomes of 
    males and females using the `inference` function, and interpret this
    interval. 

```{r echo = FALSE}
inference(y = income, x = gender, data = acs, 
          type = "ci", method = "theoretical", 
          statistic = "mean")
```

1.  Based on this interval is there a statistically significant difference between the 
    average incomes of men and women? Why, or why not?
    
1.  What is the significance level for the equivalent hypthesis test that evaluates whether 
    there is a significant difference between average incomes of men and women.

1.  Conduct this hypothesis test using the `inference` function, and interpret your results 
    in context of the data and the research question. Do your results from the
    confidence interval and the hypothesis test agree?

```{r echo = FALSE}
inference(y = income, x = gender, data = acs, type = "ht", method = "theoretical", 
          statistic = "mean", null = 0, alternative = "twosided")
```

### Confounding variables

There is a clear difference between the average salaries of men and women, but could some,
or all, of this difference be attributed to a variable other than gender? Remember that we 
call such variables confounding variables. We will evaluate whether `hrs_work` is a 
confounder for the relationship between gender and income. But before we do that, let's
first convert the `hrs_work` variable to a categorical variable (with levels `"full time"`
or `"part time"`) so that we can use methods we have learned so far in the course to run
the analysis. (Later in the course we will learn how to work with numerical explanatory
variables in a regression model setting.)

### Recoding variables

We want to create a new variable, say `emp_type`, with levels `"full time"` or `"part time"` 
depending on whether the employee works 40 hours or more per week or less than 40 hours,
respectively. We create a new variable with the `mutate` function.

```{r}
acs_emp <- acs_emp %>%
  mutate(emp_type = ifelse(hrs_work >= 40, "full time", "part time"))
```

The `ifelse()` function has three arguments: a logical test, return values for TRUE elements
of test, and return values for FALSE elements of test. In this case, `emp_type` will be coded 
as `"full time"` for observations where `hrs_work` is greater than or equal to 40, and as 
`"part time"` otherwise.

To find out what percent of the sample is full vs. part time, we turn to summary statistics:

```{r}
acs_emp %>%
  group_by(emp_type) %>%
  summarise(time_prop = n() / nrow(acs_emp))
```

Here we first grouped the data by the new `emp_type` variable, and then we calculated 
proportions of full and part time employees by first counting how many there are in each group 
(`n()`) and then dividing this number by the total number of rows of the dataset.

1.  Are females more heavily represented among full time employees or part time employees?
    Answer this question using summary statistics as above. *Hint: add one argument to the `group_by` function*.

```{r, echo = FALSE}
acs_emp %>%
  group_by(emp_type, gender) %>%
  summarise(n = n(), 
            time_prop = n() / nrow(acs_emp)) %>% 
  ungroup()
```

* * *

## Assignment 6

For the exercises below, create a new .Rmd file entitled `inference2_xxx.Rmd` where `xxx` is your last name. Include text (the questions, and your answers), as well as the R code needed to produce the answers. Knit the file to a word document and hand in the printed copy at the beginning of class next Wednesday. Be efficient with your printing - do not include extensive lines of preliminary code or the output of long vectors or dataframes. Only print the questions, brief answers, and the minimum amount of code / output to demonstrate that you know how to get the answer. 

-   Create two subsets of the `acs_emp` dataset: one for full time employees and one 
    for part time employees. No interpretation is needed for this question, 
    just the code is sufficient.

```{r, echo = FALSE}
acs_emp_full <- acs_emp %>% filter(emp_type == "full time")
acs_emp_part <- acs_emp %>% filter(emp_type == "part time")
```

-   Use a hypothesis test to evaluate whether there is a difference in average 
    incomes of **full time** male and female employees. If the difference is 
    significant, also include a confidence interval (at the equivalent confidence 
    level) estimating the magnitude of the average income difference.

```{r echo = FALSE}
inference(y = income, x = gender, data = acs_emp_full, type = "ht", method = "theoretical", 
          statistic = "mean", null = 0, alternative = "twosided")

inference(y = income, x = gender, data = acs_emp_full, type = "ci", method = "theoretical", 
          statistic = "mean", null = 0, alternative = "twosided")
```
    
-   Use a hypothesis test to evaluate whether there is a difference in average incomes
    of **part time** male and female employees. If the difference is significant,
    also include a confidence interval (at the equivalent confidence level) estimating
    the magnitude of the average income difference.
   
```{r echo = FALSE}
inference(y = income, x = gender, data = acs_emp_part, type = "ht", method = "theoretical", 
          statistic = "mean", null = 0, alternative = "twosided")

inference(y = income, x = gender, data = acs_emp_part, type = "ci", method = "theoretical", 
          statistic = "mean", null = 0, alternative = "twosided")
```   
    
-   What do your findings from these hypothesis test suggest about whether or not working
    full or part time might be a confounding variable in the relationship between
    gender and income?
    
-   What type of a test would we use to compare the average salaries across the various 
    race / ethnicity groups in this dataset? Explain your reasoning.

```{r, echo = FALSE}
acs_emp %>% count(race) # will use ANOVA
```

-   Conduct this hypothesis test using the `inference` function. Note that the response variable 
    is `income` and the explanatory variable is `race`. You will need to figure out the remaining
    arguments for the function. Use a trial-and-error approach, and let the errors inform you as
    to what else needs to be specified, and how. (Note: Use the dataset containing records from
    all employes participants: `acs_emp`) Write your hypotheses, and interpret your conclusion in
    context of the data and the research question. Note that the `inference` function by default
    uses a significance level of 0.05 for the ANOVA, and will run pairwise t-tests and report
    p-values for them if the ANOVA is significant. (Note also that you can change the significance
    level by setting `sig_level` equal to some other value.)

```{r echo = FALSE}
inference(y = income, x = race, data = acs_emp, type = "ht", method = "theoretical", 
          statistic = "mean", alternative = "greater")
```

-   Pick **another** numerical variable from the dataset to be your response variable, and 
    also pick a categorical explanatory variable (can be one we used before). Conduct the
    appropriate hypothesis test, using the `inference` function, to compare means of the
    response variable across levels of the explanatory variable. Make sure to state your 
    research question, and interpret your conclusion in context of the dataset. Note that you
    can use the complete `acs` dataset, the subsetted `acs_emp` dataset, or another subset 
    that you create.


<div id="license">
This lab was adapted from a lab written by by [Mine Cetinkaya-Rundel](https://stat.duke.edu/~mc301) and is licensed under a [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](http://creativecommons.org/licenses/by-nc-sa/4.0/).
</div>
